A Mountaintop View Requires Minimal Sorting: \break An Instance Optimal Contour Tree Algorithm

### 2014-10-27 paint-contour-tree.tex
updates to section 5

\newcommand{\LL}{\mathbb{L}}
\newcommand{\MM}{\mathbb{M}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\SSS}{\mathbb{S}}

@@

\newcommand{\build}{{\tt build}}
\newcommand{\col}{col}
\newcommand{\fcol}{full}
\newcommand{\cur}{cur}
\newcommand{\cut}{{\tt cut}}
\newcommand{\cont}{\psi}
\newcommand{\cost}{\mathop{cost}}
\newcommand{\h}{att}
\newcommand{\lift}{{\tt lift}}
\newcommand{\mcol}{mcol}
\newcommand{\merge}{{\tt merge}}
\newcommand{\onestep}{{\tt piece}}
\newcommand{\paint}{{\tt paint}}
\newcommand{\pal}{P}
\newcommand{\pmax}{P_{\max}}
\newcommand{\rain}{{\tt rain}}
\newcommand{\reeb}{\cC}
\newcommand{\redH}{\widetilde{H}}
\newcommand{\rep}{rep}
\newcommand{\st}{K}
\newcommand{\surgery}{{\tt surgery}}
\newcommand{\touch}{T}
\newcommand{\update}{{\tt update}}
\newcommand{\wet}{{\tt wet}}

@@

The previous section allows us to restrict attention to extremum dominant manifolds.
We will orient so that the extremum in question is always a \emph{minimum}.
We will fix such a simplicial complex $\MM$, with the dominant minimum $m^*$. 
We observe that the contours trees of minimum dominant manifolds have a relatively simple description.
This description is directly inspired by (and proven using) the \emph{join tree} of~\cite{csa-cctad-00}.
We first explain what these trees are.

For vertex $v$, we use $\MM^+_v$ to denote the simplicial complex obtained by only
keeping vertices $u$ such that $f(u) > f(v)$. Analogously, define $\MM^-_v$. Note that $\MM^+$ may contain
numerous connected components. The following definition of join and split trees were given by~\cite{csa-cctad-00}.
Conventially, all edges are directed from higher to lower function value.

\begin{definition} \label{join} The join tree $\cJ(\MM)$ of $\MM$ is built on vertex set $V(\MM)$.
The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex is a connected component of $\MM^+_v$
\emph{and} $v$ is connected to this component (in $\MM$). The split tree $\cS(\MM)$ is obtained
by looking at $\MM^-_v$ (or alternately, by taking the join tree of the inversion of $\MM$). 
\end{definition}

Note that unlike the contour tree, all vertices in $\MM$ are present in $\cJ(\MM)$ and $\cS(\MM)$. 
Some basic facts about these trees. All outdegrees in $\cJ(\MM)$ and all indegrees in $\cS(\MM)$ are at most $1$.
Indegree $2$ vertices in $\cJ(\MM)$ are joins, and outdegree $2$ vertices in $\cS(\MM)$ are splits.
Leaves in $\cJ(\MM)$ are maxima, and leaves in $\cS(\MM)$ are minima.

The key observation is that $\cS(\MM)$ is trivial for minimum dominant $\MM$.

@@

It is convenient to denote the non-dominant minima by $m_1, m_2, \ldots, m_k$, and the associated
splits (as given by \Lem{split}) by $s_1, s_2, \ldots, s_k$.

To prove our main theorem, we use the merging procedure from~\cite{csa-cctad-00} that constructs the contour tree
from the join and split trees. 
It actually construsts the \emph{augmented contour tree} $\cA(\MM)$, which
is the contour tree with regular points inserted into the edges. 
For any tree $T$ with vertex $v$, we define \emph{erasing} $v$ from $T$,
when both in and outdegrees are at most $1$. If $v$ is a leaf, just delete $v$. Otherwise, smooth $v$ out
(delete $v$ and connect its neighbors by an edge). This tree is denoted by $T \ominus v$.

@@

\begin{theorem} \label{thm:contour-tree} The contour tree $\cC(\MM)$ consists of all edges $\{(s_i, m_i)\}$
as well as: for any critical point $v$, edge $(w,v)$ where $w$ is the lowest non-minimal critical point
in a connected component of $\MM^+_v$ connected to $v$.
\end{theorem}

@@

Now we come to the main algorithmic contribution. We design a new algorithm for computing contour
trees of a minimum dominant simplicial complice $\MM$. As part of the initialization, we determine
all critical points in $\MM$. For any vertex $v$, we can look at $f(v) \pm \delta$-contours 
in $f|_{B_\eps(v)}$ to determine the up and down degrees. This determines the criticality (and type)
of $v$. This requires $O(|\MM|)$ time, since every facet is looked at proportional to the number of its
vertices.

{\bf Painting:} The central tool is a notion 
of \emph{painting} of $\MM$. Initially associate a color with each maximum. Denoting the set of maxima is $X$,
this is an arbitrary bijection $\chi:X \mapsto [|X|]$. Imagine there being a large
can of paint of color $\chi(x)$ at maxima $x$. We will spill different paint from each maximum and watch it flow down.
This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
\emph{So paint only flows down edges, and it does not color the interior of facets.} Furthermore, paints
do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color. This process (and indeed the entire algorithm)
works purely on the 1-skeleton of $\MM$, which is just a graph.
%
\begin{definition} \label{def:paint} Let the 1-skeleton of $\MM$ be denoted $G(\MM) = (\cV,E)$.
A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following property. 
 Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
	consisting of edges in $E$, such that all edges along this path have the same color as $x$. (The maximum $x$
	is well-defined to be the unique maximum with the same color as $e$.)

@@

\begin{proof} Consider a single push onto the stack and the current painting. We have $c \in \col(h)$, as given in pseudocode of $\update$.
If $h \in T(\rep(c))$, then obviously the next head is higher. So assume that $h \notin T(\rep(c))$.
There is obviously an edge of color $\rep(c)$ touching $h$ connected to a component $\SS$ of $\MM^+_v$. 
From the initialization, $h \in T(\rep(c'))$, for some $c'$ touching $h$ (where $c'$ is the color
of some edge connecting $h$ to $\SS$).
Applying \Lem{full},  either all edges in $\SS$ are colored $\rep(c)$ or 
there exists critical vertex $w \in \SS$ fully touched by $\rep(c)$ and another color.
In the former case, all edges connecting $\SS$ to $h$ are also colored $\rep(c)$
(since this is a painting). Hence, $\rep(c') = \rep(c)$ and $h \in T(\rep(c))$.
As we assumed this is not the case, we conclude the existence of critical vertex $w$.
Since $w$ is fully touched by $\rep(c)$, $w \in T(\rep(c))$. Since $w$
is touched by another color, $w$ has not been processed. So the highest of $T(\rep(c))$
is higher than $h$, as desired.
\ignore{
If all edges touching $h$ that connect to this component are also colored $\rep(c)$, then
In Case (i), the increase in height is obvious.
Consider Case (ii). There is a vertex $v$ touched by color $c \notin \fcol(v)$. Consider the highest
vertex $w$ that is touched by $c$ and some other color. Clearly such a vertex exists, since the condition
is also satisfied by $v$. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
and intersect some $c$-colored edge incident to $w$. Note that $\phi$ is also colored $c$,
since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
edges from the maximum to $w$. For any point on this path, the contour passing through this point must
be colored $c$.) Hence, $c \in \fcol(w)$ and $w \neq v$. Furthermore, $w$ is a critical point,
since $w$ is touched by another color. The corresponding edge cannot intersect $\phi$, so $w$
must have up-degree $2$.
All in all, when a new vertex is pushed onto the stack, it is always higher than the previous head.
}
\end{proof}

@@

\begin{proof} \Sesh{proof is a little sketchy.}
Suppose $v$ is ripe. For any $c$ touching $v$, $v$ is the highest in $T(\rep(c))$.
Consider some such $\rep(c)$, for an edge connecting to component $\SS$ of $\MM^+_v$.
Either all edges in $\SS$ are colored $\rep(c)$ or there exists critical vertex $w \in \SS$
fully touched by $\rep(c)$ and another color. In the latter case, $w$ has not been processed,
so $w \in T(\rep(c))$ (contradiction). Therefore, all edges in $\SS$ are colored $\rep(c)$.
Suppose there is some critical vertex in $\SS$. If it is not processed yet, it will
be in $T(\rep(c))$, which is not possible. Thus, all critical vertices in $\SS$ have been processed.

Consider $w$, the lowest critical vertex in $\SS$. When $w$ was processed, it was set to $\h(c')$, for
some color $c'$. At this stage, all edges in $\SS$ above $w$ were also colored $c'$ (just apply the argument
above for $w$). Since there are no critical points below $w$, all edges in $\SS$ below $w$
are also colored $c'$ (the painting property). This means that the edges connecting $\SS$
are also colored $c'$. So $c' = \rep(c)$, and $\h(\rep(c)) = w$ is a parent of $v$.

This argument can be repeated for all such components $\SS$ to complete the proof.
\end{proof}

>>>>

\newcommand{\LL}{\mathbb{L}}
\newcommand{\MM}{\mathbb{M}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\SSS}{\mathbb{S}}

@@

\newcommand{\build}{{\tt build}}
\newcommand{\col}{col}
\newcommand{\fcol}{full}
\newcommand{\cur}{cur}
\newcommand{\cut}{{\tt cut}}
\newcommand{\cont}{\psi}
\newcommand{\cost}{\mathop{cost}}
\newcommand{\h}{att}
\newcommand{\jc}{\cJ_C}
\newcommand{\lift}{{\tt lift}}
\newcommand{\mcol}{mcol}
\newcommand{\merge}{{\tt merge}}
\newcommand{\onestep}{{\tt piece}}
\newcommand{\paint}{{\tt paint}}
\newcommand{\pal}{P}
\newcommand{\pmax}{P_{\max}}
\newcommand{\rain}{{\tt rain}}
\newcommand{\reeb}{\cC}
\newcommand{\redH}{\widetilde{H}}
\newcommand{\rep}{rep}
\newcommand{\st}{K}
\newcommand{\surgery}{{\tt surgery}}
\newcommand{\touch}{T}
\newcommand{\update}{{\tt update}}
\newcommand{\wet}{{\tt wet}}

@@

The previous section allows us to restrict attention to extremum dominant manifolds.
We will orient so that the extremum in question is always a \emph{minimum}.
We will fix such a simplicial complex $\MM$, with the dominant minimum $m^*$. 
The set of non-dominant minima is denoted by $M$.
For vertex $v$, we use $\MM^+_v$ to denote the simplicial complex obtained by only
keeping vertices $u$ such that $f(u) > f(v)$. Analogously, define $\MM^-_v$. Note that $\MM^+_v$ may contain
numerous connected components.

The main theorem of this section asserts that the contours trees of minimum dominant manifolds have a simple description.
This description is based on \emph{join} and \emph{split} trees, as given by~\cite{csa-cctad-00}.
Conventially, all edges are directed from higher to lower function value.
We first explain what these trees are.

\begin{definition} \label{join} The join tree $\cJ(\MM)$ of $\MM$ is built on vertex set $V(\MM)$.
The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in a connected component of $\MM^+_v$
\emph{and} $v$ is connected to this component (in $\MM$). The split tree $\cS(\MM)$ is obtained
by looking at $\MM^-_v$ (or alternately, by taking the join tree of the inversion of $\MM$).

The critical join tree $\jc(\MM)$ is built on $V(\MM) \setminus M$. 
The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $V(\MM) \setminus M$ in a connected component of $\MM^+_v$
and $v$ is connected to this component (in $\MM$). 
\end{definition}

Some basic facts about these trees. 
%Note that unlike the contour tree, all vertices in $\MM$ are present in $\cJ(\MM)$ and $\cS(\MM)$. 
All outdegrees in $\cJ(\MM)$ and all indegrees in $\cS(\MM)$ are at most $1$.
Indegree $2$ vertices in $\cJ(\MM)$ are joins, and outdegree $2$ vertices in $\cS(\MM)$ are splits.
Leaves in $\cJ(\MM)$ are maxima, and leaves in $\cS(\MM)$ are minima.
The key observation is that $\cS(\MM)$ is trivial for minimum dominant $\MM$.

@@

To prove our main theorem, we use the merging procedure from~\cite{csa-cctad-00} that constructs the contour tree
from the join and split trees. 
It actually construsts the \emph{augmented contour tree} $\cA(\MM)$, which
is the contour tree with regular points inserted into the edges. 
Consider tree $T$ with vertex $v$ of in and outdegree at most $1$.
\emph{Erasing} $v$ from $T$ is the following operation: if $v$ is a leaf, just delete $v$. Otherwise, smooth $v$ out
(delete $v$ and connect its neighbors by an edge). This tree is denoted by $T \ominus v$.

@@

\begin{theorem} \label{thm:contour-tree} Let $\MM$ have a dominant minimum. 
The contour tree $\cC(\MM)$ consists of all edges $\{(s_i, m_i)\}$ and $\jc(\MM)$.
\end{theorem}

@@

The main algorithmic contribution is a new algorithm for computing join trees of any triangulated simplicial
complex $\MM$.

As part of the initialization, we determine
all critical points in $\MM$. For any vertex $v$, we can look at $f(v) \pm \delta$-contours 
in $f|_{B_\eps(v)}$ to determine the up and down degrees. This determines the criticality (and type)
of $v$. This requires $O(|\MM|)$ time, since every facet is looked at proportional to the number of its
vertices.

{\bf Painting:} The central tool is a notion 
of \emph{painting} of $\MM$. Initially associate a color with each maximum. Imagine there being a large
can of paint of a distinct color at each maxima $x$. We will spill different paint from each maximum and watch it flow down.
This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
\emph{So paint only flows down edges, and it does not color the interior of facets.} Furthermore, paints
do not mix, so every edge of $\MM$ gets a unique color. This process (and indeed the entire algorithm)
works purely on the 1-skeleton of $\MM$, which is just a graph.
%
\begin{definition} \label{def:paint} Let the 1-skeleton of $\MM$ have edge set $E$ and maxima $X$.
A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following property. 
 Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
	consisting of edges in $E$, such that all edges along this path have the same color as $x$. (The maximum $x$
	is well-defined to be the unique maximum with the same color as $e$.)

@@

\begin{proof} Consider a single push onto the stack and the current painting. We have $c \in \col(h)$, as given in pseudocode of $\update$.
If $h \in T(\rep(c))$, then obviously the next head is higher. So assume that $h \notin T(\rep(c))$.
There is obviously an edge of color $\rep(c)$ touching $h$ connected to a component $\PP$ of $\MM^+_v$. 
From the initialization, $h \in T(\rep(c'))$, for some $c'$ touching $h$ (where $c'$ is the color
of some edge connecting $h$ to $\PP$).
Applying \Lem{full},  either all edges in $\PP$ are colored $\rep(c)$ or 
there exists critical vertex $w \in \PP$ fully touched by $\rep(c)$ and another color.
In the former case, all edges connecting $\PP$ to $h$ are also colored $\rep(c)$
(since this is a painting). Hence, $\rep(c') = \rep(c)$ and $h \in T(\rep(c))$.
As we assumed this is not the case, we conclude the existence of critical vertex $w$.
Since $w$ is fully touched by $\rep(c)$, $w \in T(\rep(c))$. Since $w$
is touched by another color, $w$ has not been processed. So the highest of $T(\rep(c))$
is higher than $h$, as desired.
\ignore{
If all edges touching $h$ that connect to this component are also colored $\rep(c)$, then
In Case (i), the increase in height is obvious.
Consider Case (ii). There is a vertex $v$ touched by color $c \notin \fcol(v)$. Consider the highest
vertex $w$ that is touched by $c$ and some other color. Clearly such a vertex exists, since the condition
is also satisfied by $v$. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
and intersect some $c$-colored edge incident to $w$. Note that $\phi$ is also colored $c$,
since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
edges from the maximum to $w$. For any point on this path, the contour passing through this point must
be colored $c$.) Hence, $c \in \fcol(w)$ and $w \neq v$. Furthermore, $w$ is a critical point,
since $w$ is touched by another color. The corresponding edge cannot intersect $\phi$, so $w$
must have up-degree $2$.
All in all, when a new vertex is pushed onto the stack, it is always higher than the previous head.
}
\end{proof}

@@

\begin{proof} \Sesh{proof is a little sketchy.}
Suppose $v$ is ripe. For any $c$ touching $v$, $v$ is the highest in $T(\rep(c))$.
Consider some such $\rep(c)$, for an edge connecting to component $\PP$ of $\MM^+_v$.
Either all edges in $\PP$ are colored $\rep(c)$ or there exists critical vertex $w \in \PP$
fully touched by $\rep(c)$ and another color. In the latter case, $w$ has not been processed,
so $w \in T(\rep(c))$ (contradiction). Therefore, all edges in $\PP$ are colored $\rep(c)$.
Suppose there is some critical vertex in $\PP$. If it is not processed yet, it will
be in $T(\rep(c))$, which is not possible. Thus, all critical vertices in $\PP$ have been processed.

Consider $w$, the lowest critical vertex in $\PP$. Note that $w$ is a parent of $v$. When $w$ was processed, it was set to $\h(c')$, for
some color $c'$. At this stage, all edges in $\PP$ above $w$ were also colored $c'$ (just apply the argument
above for $w$). Since there are no critical points below $w$, all edges in $\PP$ below $w$
are also colored $c'$ (the painting property). This means that the edges connecting $\PP$
are also colored $c'$. So $c' = \rep(c)$, and $\h(\rep(c)) = w$ is a parent of $v$.

This argument can be repeated for all such components $\PP$ to complete the proof.
\end{proof}
